---
title: "How to use the HOQCwfs package"
output: 
  rmarkdown::html_vignette:
    toc: true 
    toc_depth: 5
vignette: >
  %\VignetteIndexEntry{How to use the HOQCwfs package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,echo=F}
```

### Introduction {#introduction}
This package contains functions to retrieve information from a Web Feature Service (WFS) resource. 
It generates HTML requests for the WFS API. 

### Workflow {#workflow}

A typical workflow looks like this :

#### Initialize {#initialize}
Load the package and indicate which WFS resource you want to access and the version that you want to use for the requests and their answers.
In this way you can avoid specifying this in the separate function calls.  
ALso specify the separator that by default will be used in generated XML fragments.

```{r initialize1}
library(HOQCwfs)
WFS_set_url()      # default https://geoweb.amstelveen.nl/geoserver/topp/wfs will be used
WFS_set_version()  # default 1.1.0 will be used
WFS_set_sep()      # default '\n' will be used
```

#### Look at the possibilities the WFS resource has to offer {#look-at-possibilities}
Retrieve the GetCapabilities document that describes the various features of the WFS resource and the actions that can be done. Because I have only read-access to WFS resources I am only interested in the query actions. An 'impression' of the rather extensive GetCapabilities xml document is given below.
```{r getcapabilities1}
xmlcap <- WFS_getcapabilities()  
class(xmlcap)
```
```{r getcapabilities2,echo=F}
HOQCutil::cap.out(print(xmlcap),width=110)
```
The xml document can be written to a local dataset with the `out_path` argument:
```{r getcapabilities3,eval=F}
xmlcap <- WFS_getcapabilities(out_path='mycap.xml') 
```
[[back to contents]](#TOC)
 
#### Look at the various features of the WFS resource {#look-at-features}
From the GetCapabilities document we can extract the features that are available:
```{r featuretypes1}
ft1 <- WFS_featuretypes(xmlcap)
dim(ft1)
```
```{r featuretypes2,echo=F}
ft2 <- WFS_featuretypes(xmlcap,
       filternames=stringr::fixed("bomen", ignore_case = T))
```
When there are many features (`r dim(ft1)[1]` in this case) and we are just interested in feature names that contain the characters 'bomen' (i.e. 'trees') we can filter the names in the following way that results in only `r dim(ft2)[1]` cases
```{r ref.label='featuretypes2',eval=F}
```
```{r featuretypes3}
class(ft2)
dim(ft2)
ft2$layer
str(ft2[3,]) # show only featuretype 'topp:gidw_groenbomen' 
```
```{r describefeaturetype1,echo=F}
bomen_fields_1_3<- WFS_describefeaturetype(ft2$layer[c(1,3)])
```
[[back to contents]](#TOC)

#### Look at the data fields of featuretypes {#look-at-data-fields}
For this we use the function `WFS_describefeaturetype` that executes the WFS `DescribeFeatureType` request. 
As an example lets us look at the first and third featuretype in ft2 (`r glue::glue_collapse(glue::backtick(ft2$layer[c(1,3)]),last=" and ")`). Here we only show the first `5` rows of the result tibble of `r dim(bomen_fields_1_3)[1]` rows.
```{r ref.label='describefeaturetype1',eval=F}
```
```{r describefeaturetype2}
head(bomen_fields_1_3,5)
```
Because we later will use the fields of `topp:gidw_groenbomen` we will show all fields for this featuretype. Because we now request only one featuretype we can save the result of `DescribeFeatureType` in an xml file. Then we have the standard output:
```{r describefeaturetype3}
WFS_describefeaturetype("topp:gidw_groenbomen",out_path = 'desc.xml')
```
and the `desc.xml` file with its contents:
```{r describefeaturetype4,echo=F}
HOQCutil::cap.out(cat(readLines("desc.xml",warn = F),sep='\n'), width=110) 
```
```{r unlink, echo=F}
unlink('desc.xml')
```
[[back to contents]](#TOC)

#### Retrieve the actual data {#retrieve}
Now that we know the metadata of our resource we can actually retrieve the attributes and geometrical data of the features. We do this with `WFS_getfeature` to:  

- [retrieve all features with all attributes and the geometry](#retrieve-all): The `id` attribute will also be returned 
- [retrieve some features](#retrieve-some-features) and the geometry: add `cql_filter=` or `filter=` with a filter expression
- [retrieve all features for some attributes](#retrieve-attributes) and the geometry: add `propertyname='a1,a2'` to select the attributes `a1` and `a2`. The `id` attribute and the `geometry` will also be returned 
- [change the Spatial Reference System](#change-crs) (i.e. the coordinates of the features)
- [retrieve a subset](#retrieve-subset) of one of the above: add `maxfeatures=` or `count=` optionally in combination with `startindex=`
- [retrieve all features in a bounding box](#retrieve_boundingbox)
- [retrieve features in a bounding box with additional filtering](#retrieve_boundingbox-cqlfilter)

If we are only interested in the [number of features](#number-of-features) that would be returned we can add `resultType=hits` to any of the above. This only returns this number without any other data.

Examples:

[[back to contents]](#TOC) 

##### retrieve all features with all attributes and the geometry {#retrieve-all}
```{r retrieve-all1,eval=F}
typename  <- 'topp:gidw_groenbomen'
wfs1      <-  WFS_getfeature(typename)
class(wfs1)
print(wfs1,n=5) # show only first 5 features
```
```{r retrieve-all2,echo=F}
typename  <- 'topp:gidw_groenbomen'
```
TODO remove eval=F  

[[back to contents]](#TOC) [[back to 'Retrieve the actual data']](#retrieve)

##### retrieve some features {#retrieve-some-features}
For non-complex queries the `cql_filter` can be used. The features for the tree `Prunus serrulata 'Kanzan'` can be queried with the following code. The number of quotes in a query is often confusing (for me at least): here a set of quotes is needed because `boom_omschrijf` is a character field and quotes around 'Kanzan' have to be duplicated. 

```{r retrieve-some-features}
tree_name <- "Prunus serrulata ''Kanzan''"  
tree_exp  <- glue::glue("boom_omschrijf='{tree_name}'")
print(tree_exp)
wfs2      <-  WFS_getfeature(typename,cql_filter=tree_exp )
print(wfs2,n=5,digits=5) # show only first 5 features, 5 digits in geometry
```
[[back to contents]](#TOC) [[back to 'Retrieve the actual data']](#retrieve)

##### retrieve some attributes  {#retrieve-attributes}
When we are only interested in the 'omschrijvingen' (descriptions) fields we can add the `propertyname` argument to request the fields `boom_omschrijf` and `objec_omschrijf`. Note that the `id` and `geometry` fields are always included.

```{r retrieve-attributes}
tree_name <- "Prunus serrulata ''Kanzan''"  
tree_exp  <- glue::glue("boom_omschrijf='{tree_name}'")
fields    <- 'boom_omschrijf,objec_omschrijf'
wfs3      <-  WFS_getfeature(typename,cql_filter=tree_exp,propertyname=fields )
print(wfs3,n=5,digits=5) # show only first 5 features, 5 digits in geometry
```
[[back to contents]](#TOC) [[back to 'Retrieve the actual data']](#retrieve)

##### retrieve the number of results {#number-of-features}
If we want to know beforehand the number of features that a query would return we can add the `resultType=hits` argument:

```{r number-of-features}
  WFS_getfeature(typename,cql_filter=tree_exp,propertyname=fields,resultType='hits' )
```
[[back to contents]](#TOC) [[back to 'Retrieve the actual data']](#retrieve)

##### change the Spatial Reference System  {#change-crs}
In the output of the previous examples we saw that the coordinates in the geometry were denoted in the Coordinate Reference System (CRS) that is used in the Netherlands: 'Amersfoort / RD New' also known as 'EPSG:28992'. If we want to have these coordinates denoted in the more commonly used 'WGS 84' (EPSG:4326) in terms of longitude and latitude we can add the `srsName` argument. SRS ([Spatial Reference System](https://en.wikipedia.org/wiki/Spatial_reference_system)) is a synonym for CRS.

```{r change-crs}
wfs4      <-  WFS_getfeature(typename,cql_filter=tree_exp,propertyname=fields,srsName='EPSG:4326' )
print(wfs4,n=5,digits=5) # show only first 5 features, 5 digits in geometry
```
[[back to contents]](#TOC) [[back to 'Retrieve the actual data']](#retrieve)

##### retrieve a subset of the query results with `startindex`, `maxfeatures` and `count` and using `verbose` {#retrieve-subset}
With the `startindex` and `maxfeatures` (1.1.0) or `count` (2.0.0) argument a subset of a query is produced. The `WFS__getfeature` function will use the correct version. By adding the `verbose=T` clause the generated request url will be displayed so that you can verify this. Notice that `startIndex=2` indicates that the first feature to be retrieved is number 3. In other words `startIndex` indicates the number of records to skip.

```{r retrieve-subset1,eval=F}
wfs5a     <-  WFS_getfeature(typename,cql_filter=tree_exp,startIndex=2,maxfeatures=6,
                             propertyname=fields,srsName='EPSG:4326'
                             ,verbose=T,version='1.1.0' )
```
```{r retrieve-subset2,echo=F}

HOQCutil::cap.out({
wfs5a     <-  WFS_getfeature(typename,cql_filter=tree_exp,startIndex=2,maxfeatures=6,
                             propertyname=fields,srsName='EPSG:4326'
                             ,verbose=T,version='1.1.0' )  
}, width=93)
```
```{rretrieve-subset3}
print(wfs5a,n=5,digits=5) # show only first 5 features, 5 digits in geometry
```

```{r retrieve-subset4,eval=F}
wfs5b     <-  WFS_getfeature(typename,cql_filter=tree_exp,startIndex=2,maxfeatures=6,
                             propertyname=fields,srsName='EPSG:4326'
                             ,verbose=T,version='2.0.0' )
```
```{r retrieve-subset5,echo=F}
HOQCutil::cap.out({
wfs5b     <-  WFS_getfeature(typename,cql_filter=tree_exp,startIndex=2,maxfeatures=6,
                             propertyname=fields,srsName='EPSG:4326'
                             ,verbose=T,version='2.0.0' )
}, width=93)
```
```{r retrieve-subset6}
identical(wfs5a,wfs5b)
```
[[back to contents]](#TOC) [[back to 'Retrieve the actual data']](#retrieve)

##### retrieve features in a bounding box {#retrieve_boundingbox}
When you want to retrieve all features in a bounding box the `bbox=` argument can be used. When specifying the coordinates of the bounding box it is assumed that the default CRS is used.
The default CRS for `typename` is `EPSG:28992` (see [Look at the various features of the WFS resource](#look-at-features) ). Therefore when using `WGS 84` (`EPSG:4326`) the EPSG code has to be added to the specification. Note that this only impacts the features that are selected and not their coordinates. Use `srsName='EPSG:4326'` when you want `WGS 84` coordinates.  
In the two examples below we are interested in the same area but in the first case we use `EPSG:28992` coordinates and in the second `WGS 84` coordinates. Therefore the same features are selected
```{r retrieve_boundingbox1}
bbox_wgs84 <- c(4.860793, 52.313319, 4.861587, 52.316493 )
bbox_28992 <- convert_bbox(bbox_wgs84,'EPSG:4326','EPSG:28992')  

(bbox_spec <- glue::glue_collapse(bbox_28992,sep=','))
wfs6a      <-  WFS_getfeature(typename,bbox=bbox_spec,
                             propertyname=fields )
print(wfs6a,n=5,digits=5)

(bbox_spec <- paste0(glue::glue_collapse(bbox_wgs84,sep=','),',EPSG:4326'))
wfs6b      <-  WFS_getfeature(typename,bbox=bbox_spec,
                             propertyname=fields )
print(wfs6b,n=5,digits=5)
```
[[back to contents]](#TOC) [[back to 'Retrieve the actual data']](#retrieve)

##### retrieve features in a bounding box with additional filtering {#retrieve_boundingbox-cqlfilter}
A combination of `cql_filter=` or `filter=` with `bbox=` is not allowed. In that case the bbox filtering has to be included in the `cql_filter=` or `filter=` construct. E.g. (and note the quotes around `EPSG:4326` where these were not necessary in the `bbox=` case) :

```{r retrieve_boundingbox-cqlfilter}
tree_name <- "Alnus glutinosa ''Laciniata''"  
tree_exp  <- glue::glue("boom_omschrijf='{tree_name}'")
bbox_spec <- glue::glue_collapse(bbox_28992,sep=',')
( cql_spec<- paste0(tree_exp,' and bbox(geometrie,',bbox_spec,')') )
wfs7a      <-  WFS_getfeature(typename,cql_filter=cql_spec,
                             propertyname=fields )
print(wfs7a,n=5,digits=5)

bbox_spec <- paste0(glue::glue_collapse(bbox_wgs84,sep=','),",'EPSG:4326'")
( cql_spec<- paste0(tree_exp,' and bbox(geometrie,',bbox_spec,')') )
wfs7b      <-  WFS_getfeature(typename,cql_filter=cql_spec,
                             propertyname=fields,srsName='EPSG:4326' )
print(wfs7b,n=5,digits=5)
```

[[back to contents]](#TOC) [[back to 'Retrieve the actual data']](#retrieve)


